---
title: "Supervised Learning"
author: "Mandip Chaudhary (17)"
format: pptx
editor: visual
---

## Introduction

-   Dataset: Motor Trend Car Road Tests (mtcars)
-   Goal: Predict car fuel efficiency (MPG)
-   Methods used: Simple Linear Regression, Multiple Regression, LASSO
-   Focus on model quality and prediction capability

## Data Preparation {.smaller}

```{r}
#| echo: false
#| warning: false
library(datasets)
set.seed(17)
train_index <- sample(2, nrow(mtcars), replace = T, prob = c(0.7, 0.3))
train_data <- mtcars[train_index==1, ]
test_data <- mtcars[train_index==2, ]
```

::::: columns
::: {.column width="50%"}
**Training Data (70%)**

-   Used to build models
-   Size: `r nrow(train_data)` cars
:::

::: {.column width="50%"}
**Testing Data (30%)**

-   Used to evaluate models
-   Size: `r nrow(test_data)` cars
:::
:::::

## Simple Linear Regression Results

**Individual Models: MPG vs Each Variable**

Top performers (R² values):

```{r}
#| echo: false
predictors <- setdiff(names(train_data), "mpg")
r_squared <- sapply(predictors, function(pred) {
  model <- lm(paste("mpg ~", pred), data = train_data)
  summary(model)$r.squared
})
top_3 <- head(sort(r_squared, decreasing = TRUE), 3)
for(name in names(top_3)) {
  cat("- ", name, ": ", round(top_3[name]*100, 1), "%\n", sep="")
}
```

## Why Aren't These Models BLUE?

BLUE = Best Linear Unbiased Estimator

Common violations found:

1.  Non-linear relationships with MPG
2.  Uneven spread in residuals
3.  Discrete predictor issues
4.  Zero mean assumption violations

## Multiple Regression Approach

1.  Started with significant predictors (p \< 0.05)
2.  Checked for multicollinearity using VIF
3.  Removed variables with VIF ≥ 10
4.  Created final multiple regression model

## LASSO Regression {.smaller}

**What is LASSO?**

-   Least Absolute Shrinkage and Selection Operator
-   Automatically selects important variables
-   Helps with multicollinearity
-   Can improve prediction accuracy

```{r}
#| echo: false
#| warning: false
library(glmnet)
x_train <- as.matrix(train_data[, predictors])
y_train <- train_data$mpg
lasso_mod <- cv.glmnet(x_train, y_train, alpha=1)
```

## Model Performance Comparison

```{r}
#| echo: false
# Function to get metrics
get_metrics <- function(actual, predicted) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae <- mean(abs(actual - predicted))
  r2 <- 1 - sum((actual - predicted)^2)/sum((actual - mean(actual))^2)
  c(RMSE = rmse, MAE = mae, R2 = r2)
}

# Multiple regression metrics
multi_model <- lm(mpg ~ wt + hp + am, data=train_data)  # Example variables
multi_train_pred <- predict(multi_model, train_data)
multi_test_pred <- predict(multi_model, test_data)
multi_train_metrics <- get_metrics(train_data$mpg, multi_train_pred)
multi_test_metrics <- get_metrics(test_data$mpg, multi_test_pred)

# LASSO metrics
lasso_train_pred <- predict(lasso_mod, x_train)
x_test <- as.matrix(test_data[, predictors])
lasso_test_pred <- predict(lasso_mod, x_test)
lasso_train_metrics <- get_metrics(train_data$mpg, lasso_train_pred)
lasso_test_metrics <- get_metrics(test_data$mpg, lasso_test_pred)

# Create comparison table
metrics <- rbind(
  "Multiple (Train)" = round(multi_train_metrics, 2),
  "Multiple (Test)" = round(multi_test_metrics, 2),
  "LASSO (Train)" = round(lasso_train_metrics, 2),
  "LASSO (Test)" = round(lasso_test_metrics, 2)
)
print(metrics)
```

## Can We Use This Model for Prediction?

**Yes, but with cautions:**

1.  Works best within training data range
2.  Avoid extrapolation (predictions outside data range)
3.  Consider prediction intervals for uncertainty
4.  Regular model updates recommended

## Key Takeaways

1.  Data splitting is crucial for model validation
2.  Simple models weren't BLUE - assumptions violated
3.  LASSO helped handle multicollinearity
4.  Model works but has clear limitations
5.  Always validate assumptions

## Recommendations

1.  Gather more modern car data
2.  Consider non-linear relationships
3.  Include more relevant features
4.  Use prediction intervals
5.  Regular model updates

## Questions?

Thank you for your attention!

Contact: Mandip Chaudhary
