---
title: "Understanding Car Fuel Efficiency"
subtitle: "A Data-Driven Analysis Using the mtcars Dataset"
author: "Mandip Chaudhary (17)"
format: pptx
editor: visual
---

## Understanding Our Data {.larger}

### About the Car Dataset

**What We're Working With:** - 32 different car models from 1973-74 - Focus: How car features affect fuel efficiency - Goal: Predict miles per gallon (MPG)

**Key Car Features:** - **Engine Stats:** \* Number of cylinders (4, 6, or 8) \* Engine size (displacement) \* Horsepower - **Physical Features:** \* Car weight (in 1000 lbs) \* Transmission type (Auto/Manual)

```{r}
#| echo: false
#| warning: false
library(datasets)
library(corrplot)
# Create correlation matrix visualization
correlation_matrix <- cor(mtcars)
corrplot(correlation_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Variables")
```

## Data Preparation: Training vs Testing Split {.larger}

### Robust Split Methodology

```{r}
#| echo: true
#| warning: false
#| message: false
library(datasets)
library(glmnet)
set.seed(17)

# Data preparation
train_index <- sample(2, nrow(mtcars), replace = T, prob = c(0.7, 0.3))
train_data <- mtcars[train_index==1, ]
test_data <- mtcars[train_index==2, ]

# Ensure train_data is properly defined in global environment
assign("train_data", train_data, envir = .GlobalEnv)
assign("test_data", test_data, envir = .GlobalEnv)

# Prepare LASSO matrices once
predictors <- setdiff(names(train_data), "mpg")
x_train <- as.matrix(train_data[, predictors])
y_train <- train_data$mpg

# Define these in global environment for later use
assign("x_train", x_train, envir = .GlobalEnv)
assign("y_train", y_train, envir = .GlobalEnv)
```

**Training Dataset (70%)** - Size: `r nrow(train_data)` observations - Purpose: Model building and parameter estimation - Used for: Initial modeling, BLUE analysis, VIF calculations

**Testing Dataset (30%)** - Size: `r nrow(test_data)` observations - Purpose: Unbiased performance evaluation - Used for: Model validation, prediction accuracy

```{r}
#| echo: false
# Create summary statistics table
train_summary <- summary(train_data)
test_summary <- summary(test_data)
```

### Data Characteristics:

Training Set MPG Statistics: - Mean: `r round(mean(train_data$mpg), 2)` - Median: `r round(median(train_data$mpg), 2)` - Range: `r round(min(train_data$mpg), 2)` - `r round(max(train_data$mpg), 2)`

## Simple Linear Regression Analysis {.larger}

### Individual Variable Performance

```{r}
#| echo: false
#| warning: false
# Calculate R-squared and p-values for each predictor
predictors <- setdiff(names(train_data), "mpg")
results <- data.frame(
  Variable = predictors,
  R_squared = NA,
  P_value = NA
)

for(i in seq_along(predictors)) {
  model <- lm(paste("mpg ~", predictors[i]), data = train_data)
  results$R_squared[i] <- summary(model)$r.squared
  results$P_value[i] <- summary(model)$coefficients[2,4]
}

results <- results[order(-results$R_squared),]
```

**Top Performing Variables (R² values):**

```{r}
#| echo: false
for(i in 1:4) {
  cat(sprintf("%s: %.1f%% (p = %.10f)\n", 
              results$Variable[i], 
              results$R_squared[i]*100,
              results$P_value[i]))
}
```

## BLUE Test Results {.larger}

### Model Assessment Summary

```{r}
#| echo: false
library(knitr)
blue_results <- data.frame(
  Variable = c("cyl", "wt", "disp", "hp", "drat", "qsec", "vs", "am", "gear", "carb"),
  R2 = c("Yes (0.73)", "Yes (0.73)", "Yes (0.68)", "Yes (0.57)", "Yes (0.53)", 
         "No (0.23)", "No (0.36)", "No (0.41)", "No (0.19)", "No (0.34)"),
  ANOVA = c("Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)",
            "Yes (0.031)", "Yes (0.005)", "Yes (0.002)", "No (0.052)", "Yes (0.007)"),
  Intercept = c("Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)", "No (0.249)",
                "No (0.424)", "Yes (<0.001)", "Yes (<0.001)", "No (0.341)", "Yes (<0.001)"),
  Coefficient = c("Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)", "Yes (<0.001)",
                  "Yes (0.031)", "Yes (0.005)", "Yes (0.002)", "No (0.052)", "Yes (0.007)"),
  BLUE = c("Yes", "Yes", "Yes", "Yes", "No", "No", "No", "No", "No", "No")
)

kable(blue_results, 
      col.names = c("Variable", "R² > 0.5", "ANOVA sig.", "Intercept sig.", "Coef sig.", "BLUE"),
      align = c('l', 'c', 'c', 'c', 'c', 'c'),
      row.names = FALSE)
```

**BLUE Requirements:**
- R² > 0.5 indicates good model fit
- All significance tests must have p < 0.05
- Model must pass all criteria to be BLUE

**Variable Relationships:**

```{r}
#| echo: false
#| warning: false
par(mfrow=c(2,2))
for(i in 1:4) {
  plot(train_data[[results$Variable[i]]], train_data$mpg,
       xlab=results$Variable[i], ylab="MPG",
       main=paste("MPG vs", results$Variable[i]))
  abline(lm(mpg ~ get(results$Variable[i]), data=train_data), col="red")
}
```

## Most Important Car Features {.larger}

### Key Findings from Analysis

**Top 4 Most Influential Features:**

1.  **Number of Cylinders (cyl)** (R² = 73.24%)
    -   Explains 73% of MPG variation.
    -   Very strong relationship with fuel efficiency.
2.  **Car Weight (wt)** (R² = 73.16%)
    -   Almost equally important as cylinders.
    -   Heavier cars use more fuel.
3.  **Engine Size (disp)** (R² = 67.87%)
    -   Larger engines typically use more fuel.
    -   Strong predictor of MPG.
4.  **Horsepower (hp)** (R² = 56.83%)
    -   More power means more fuel consumption.
    -   Moderate to strong relationship.

**Why These Matter:** - All have strong statistical evidence (p \< 0.05). - Each explains more than 50% of MPG variation. - Show clear relationships with fuel efficiency.

## Combining Car Features {.larger}

### Building a Better Model

**Initial Approach:** - Combined all important features together - Found some features were too closely related - Needed to simplify the model

**Problem Found:** - Engine size (disp) overlapped with other features - VIF Analysis showed redundant information: \* Engine size VIF \> 10 (too high) \* Other features had acceptable VIF values

**Final Model:** - Kept three key features: 1. Car Weight (wt) 2. Horsepower (hp) 3. Transmission Type (am) - Removed redundant engine size measure. - Maintained prediction accuracy.

## LASSO Model Results {.larger}

### Smart Feature Selection

**What LASSO Does:** - Automatically picks the most important car features - Removes less important features - Helps avoid overcomplicating the model

**Key Results:**

1.  Most Important Features Selected:
    -   Car Weight (wt)
    -   Horsepower (hp)
    -   Transmission Type (Manual/Auto)
2.  Model Performance:
    -   Training RMSE: Shows typical prediction error
    -   Testing RMSE: Confirms model reliability
    -   Similar performance to standard regression

**Why This Matters:** - Confirms our earlier findings. - Simpler model, similar accuracy. - More reliable predictions.

### Visualizing LASSO: Finding the Best Balance

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "LASSO Cross-Validation: This chart helps LASSO find the best balance. It's like tuning a radio to get the clearest signal – we're finding the 'sweet spot' for our model. The dotted line shows the best point."
# Use the globally defined x_train and y_train
if (exists("x_train") && exists("y_train")) {
  set.seed(17)  # For reproducibility of cv.glmnet
  lasso_cv <- cv.glmnet(x_train, y_train, alpha=1)
  plot(lasso_cv)
  # Make lasso_cv available globally
  assign("lasso_cv", lasso_cv, envir = .GlobalEnv)
} else {
  plot(NULL, xlim=c(0,1), ylim=c(0,1), 
       xlab="log(Lambda)", ylab="Mean-Squared Error",
       main="LASSO Cross-Validation")
  text(0.5, 0.5, "Data not available for LASSO CV plot")
}
```

This plot shows how the model's error changes as LASSO adjusts its "strictness" (lambda). The vertical dotted lines indicate optimal lambda values, helping us choose the best model.

### Visualizing LASSO: Which Features Matter Most?

```{r}
#| echo: false
#| warning: false
#| fig-cap: "LASSO Coefficient Path: Imagine LASSO as a manager deciding which players (features) are most valuable. This plot shows how, as LASSO gets stricter, it keeps only the star players and lets others go. Lines that stay high are important; lines that drop to zero are removed."

library(glmnet)
library(caret)

lasso_model_plot <- train(
  mpg~ (cyl + wt + disp + hp),
  data= train_data,
  method="glmnet",
)
plot(varImp(lasso_model_plot), main = "Variable Importance from LASSO")
```

This chart shows the "MVPs" (Most Valuable Predictors) selected by LASSO. Bigger values (further from zero) mean more impact on fuel efficiency.

## Model Performance Comparison {.larger}

### Comprehensive Metrics Analysis

```{r}
#| echo: false
#| warning: false

library(glmnet)

# Prepare data for LASSO. Note that train_data, predictors, x_train, y_train, and lasso_cv
# are also defined or used in earlier chunks. This chunk redefines them for clarity
# and to ensure the metrics are based on these specific definitions.
predictors <- setdiff(names(train_data), "mpg")
x_train <- as.matrix(train_data[, predictors])
y_train <- train_data$mpg
# This lasso_cv is specific to this chunk for calculating lasso_model's lambda.
# It will be identical to the one in the plot chunk if inputs are identical.
lasso_cv_metrics <- cv.glmnet(x_train, y_train, alpha=1)

# Function for metrics
get_metrics <- function(actual, predicted) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae <- mean(abs(actual - predicted))
  r2 <- 1 - sum((actual - predicted)^2)/sum((actual - mean(actual))^2)
  c(RMSE = rmse, MAE = mae, R2 = r2)
}

# Calculate metrics for both models
final_multi_model <- lm(mpg ~ wt + hp + am, data=train_data)
# Use lambda from lasso_cv_metrics for consistency within this chunk's calculations
lasso_model <- glmnet(x_train, y_train, alpha=1, lambda=lasso_cv_metrics$lambda.min)

# Training metrics
multi_train_pred <- predict(final_multi_model, train_data)
lasso_train_pred <- predict(lasso_model, x_train)

multi_metrics <- get_metrics(train_data$mpg, multi_train_pred)
lasso_metrics <- get_metrics(train_data$mpg, lasso_train_pred)

metrics_table <- rbind(
  "Multiple Regression" = round(multi_metrics, 3),
  "LASSO Regression" = round(lasso_metrics, 3)
)

print(metrics_table)
```

```{r}
#| echo: false
#| warning: false
# Code for 6000 lbs car prediction
new_car <- data.frame(
  wt = 6,
  hp = mean(train_data$hp), # Ensure train_data is available
  am = mean(train_data$am)  # Ensure train_data is available
)
# Ensure final_multi_model is defined from the previous chunk
pred_multi <- predict(final_multi_model, new_car, interval="prediction")
```

## Predicting Fuel Efficiency {.larger}

### Example: 6000 lbs Car

**Prediction:** - For a 6000 lbs car, predicted MPG is `r round(pred_multi[1], 2)` - 95% sure it's between `r round(pred_multi[2], 2)` and `r round(pred_multi[3], 2)` MPG

**Important Note:** - This is an estimate for a very heavy car - Our data doesn't include cars this heavy - Predictions outside data range are less reliable

## Key Takeaways & Next Steps {.larger}

### Main Findings:

1.  Car Weight, Horsepower, and Transmission are key for MPG
2.  LASSO helps simplify models without losing accuracy
3.  Predictions are useful but have limits

### Recommendations:

1.  Use modern car data for better insights
2.  Explore more advanced modeling techniques
3.  Always be cautious with predictions outside data range

## Questions?

Thank you for your attention!
