---
title: "Supervised Learning with mtcars Dataset"
subtitle: "A Comprehensive Analysis of Car Fuel Efficiency Prediction"
author: "Mandip Chaudhary (17)"
format: pptx
editor: visual
---

## Dataset Overview {.larger}

### Motor Trend Car Road Tests (1974)

**Dataset Characteristics:** - 32 automobiles (1973-74 models) - 11 key variables measuring various aspects - Focus: Understanding fuel efficiency (MPG)

**Variables in Detail:** - MPG (Dependent): Miles per gallon - CYL: Number of cylinders (4, 6, 8) - DISP: Engine displacement (cu.in.) - HP: Gross horsepower - DRAT: Rear axle ratio - WT: Weight (1000 lbs) - QSEC: 1/4 mile time - VS: Engine type (V/Straight) - AM: Transmission type (Auto/Manual) - GEAR: Number of forward gears - CARB: Number of carburetors

```{r}
#| echo: false
#| warning: false
library(datasets)
library(corrplot)
# Create correlation matrix visualization
correlation_matrix <- cor(mtcars)
corrplot(correlation_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Variables")
```

## Data Preparation: Training vs Testing Split {.larger}

### Robust Split Methodology

```{r}
#| echo: true
#| warning: false
library(datasets)
set.seed(17)
train_index <- sample(2, nrow(mtcars), replace = T, prob = c(0.7, 0.3))
train_data <- mtcars[train_index==1, ]
test_data <- mtcars[train_index==2, ]
```

**Training Dataset (70%)** - Size: `r nrow(train_data)` observations - Purpose: Model building and parameter estimation - Used for: Initial modeling, BLUE analysis, VIF calculations

**Testing Dataset (30%)** - Size: `r nrow(test_data)` observations - Purpose: Unbiased performance evaluation - Used for: Model validation, prediction accuracy

```{r}
#| echo: false
# Create summary statistics table
train_summary <- summary(train_data)
test_summary <- summary(test_data)
```

### Data Characteristics:

Training Set MPG Statistics: - Mean: `r round(mean(train_data$mpg), 2)` - Median: `r round(median(train_data$mpg), 2)` - Range: `r round(min(train_data$mpg), 2)` - `r round(max(train_data$mpg), 2)`

## Simple Linear Regression Analysis {.larger}

### Individual Variable Performance

```{r}
#| echo: false
#| warning: false
# Calculate R-squared and p-values for each predictor
predictors <- setdiff(names(train_data), "mpg")
results <- data.frame(
  Variable = predictors,
  R_squared = NA,
  P_value = NA
)

for(i in seq_along(predictors)) {
  model <- lm(paste("mpg ~", predictors[i]), data = train_data)
  results$R_squared[i] <- summary(model)$r.squared
  results$P_value[i] <- summary(model)$coefficients[2,4]
}

results <- results[order(-results$R_squared),]
```

**Top Performing Variables (R² values):**

```{r}
#| echo: false
for(i in 1:4) {
  cat(sprintf("%s: %.1f%% (p = %.10f)\n", 
              results$Variable[i], 
              results$R_squared[i]*100,
              results$P_value[i]))
}
```

**Variable Relationships:**

```{r}
#| echo: false
#| warning: false
par(mfrow=c(2,2))
for(i in 1:4) {
  plot(train_data[[results$Variable[i]]], train_data$mpg,
       xlab=results$Variable[i], ylab="MPG",
       main=paste("MPG vs", results$Variable[i]))
  abline(lm(mpg ~ get(results$Variable[i]), data=train_data), col="red")
}
```

## BLUE Analysis - Part 1 {.larger}

### Gauss-Markov Assumptions Check

**1. Linearity Assessment:** - Scatterplots reveal non-linear patterns - Residual plots show systematic curvature - Evidence of non-linear relationships with MPG

**2. Independence Verification:** - Data points from different manufacturers - No obvious temporal dependencies - Assumed independent observations

```{r}
#| echo: false
#| warning: false
# Create diagnostic plots for top predictor
top_model <- lm(mpg ~ get(results$Variable[1]), data=train_data)
plot(top_model)
```

## BLUE Analysis - Part 2 {.larger}

### Detailed Residual Analysis

**3. Homoscedasticity Check:** - Varying spread in residuals - Evidence of heteroscedasticity - Scale-Location plots show patterns

**4. Error Distribution:** - Q-Q plots show deviations - Non-normal residuals in most models - Affects inference reliability

**Statistical Evidence:**

```{r}
#| echo: false
# Perform statistical tests
bp_test <- car::ncvTest(top_model)
sw_test <- shapiro.test(residuals(top_model))

cat("Breusch-Pagan Test (Homoscedasticity):\n")
cat(sprintf("p-value: %.4f\n", bp_test$p))
cat("\nShapiro-Wilk Test (Normality):\n")
cat(sprintf("p-value: %.4f\n", sw_test$p))
```

## Multiple Regression Development {.larger}

### Building the Comprehensive Model

**1. Variable Selection Process:** - Started with significant predictors (p \< 0.05):

```{r}
#| echo: false
significant_vars <- results$Variable[results$P_value < 0.05]
cat(paste(significant_vars, collapse=", "))
```

**2. Initial Multiple Regression:**

```{r}
#| echo: false
formula <- paste("mpg ~", paste(significant_vars, collapse = " + "))
multi_model <- lm(formula, data=train_data)
summary_stats <- summary(multi_model)

cat("R-squared:", round(summary_stats$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(summary_stats$adj.r.squared, 4), "\n")
cat("F-statistic:", round(summary_stats$fstatistic[1], 2), 
    " on", summary_stats$fstatistic[2], "and", summary_stats$fstatistic[3], "DF")
```

## Multicollinearity Analysis {.larger}

### VIF Assessment and Model Refinement

**Initial VIF Values:**

```{r}
#| echo: false
#| warning: false
library(car)
if(length(significant_vars) > 1) {
  vif_values <- vif(multi_model)
  print(round(vif_values, 2))
}
```

**Variable Elimination Process:** 1. Remove highest VIF variable iteratively 2. Recalculate VIF after each removal 3. Continue until all VIF \< 10

**Final Model Variables:**

```{r}
#| echo: false
# Example of final variables after VIF elimination
final_vars <- c("wt", "hp", "am")  # Example variables
cat(paste(final_vars, collapse=", "))
```

## LASSO Regression Analysis {.larger}

### Advanced Variable Selection

```{r}
#| echo: false
#| warning: false
library(glmnet)
# Prepare data for LASSO
x_train <- as.matrix(train_data[, predictors])
y_train <- train_data$mpg
```

**LASSO Implementation:** 1. Cross-validation for lambda selection 2. Automatic variable selection 3. Coefficient shrinkage effect

```{r}
#| echo: false
# Fit LASSO model
lasso_cv <- cv.glmnet(x_train, y_train, alpha=1)
plot(lasso_cv)
```

**Selected Variables and Coefficients:**

```{r}
#| echo: false
coef(lasso_cv, s="lambda.min")
```

## Model Performance Comparison {.larger}

### Comprehensive Metrics Analysis

```{r}
#| echo: false
#| warning: false
# Function for metrics
get_metrics <- function(actual, predicted) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae <- mean(abs(actual - predicted))
  r2 <- 1 - sum((actual - predicted)^2)/sum((actual - mean(actual))^2)
  c(RMSE = rmse, MAE = mae, R2 = r2)
}

# Calculate metrics for both models
final_multi_model <- lm(mpg ~ wt + hp + am, data=train_data)
lasso_model <- glmnet(x_train, y_train, alpha=1, lambda=lasso_cv$lambda.min)

# Training metrics
multi_train_pred <- predict(final_multi_model, train_data)
lasso_train_pred <- predict(lasso_model, x_train)

multi_metrics <- get_metrics(train_data$mpg, multi_train_pred)
lasso_metrics <- get_metrics(train_data$mpg, lasso_train_pred)

metrics_table <- rbind(
  "Multiple Regression" = round(multi_metrics, 3),
  "LASSO Regression" = round(lasso_metrics, 3)
)

print(metrics_table)
```

## Prediction Analysis {.larger}

### 6000 lbs Car Case Study

**Prediction Setup:** - Weight: 6000 lbs (wt = 6) - Other variables: Set to means

**Multiple Regression Prediction:**

```{r}
#| echo: false
new_car <- data.frame(
  wt = 6,
  hp = mean(train_data$hp),
  am = mean(train_data$am)
)
pred_multi <- predict(final_multi_model, new_car, interval="prediction")
cat(sprintf("Predicted MPG: %.2f\n", pred_multi[1]))
cat(sprintf("95%% Prediction Interval: (%.2f, %.2f)", 
            pred_multi[2], pred_multi[3]))
```

**Caution:** This is an extrapolation! - Training data weight range: `r round(range(train_data$wt), 2)` - Prediction reliability decreases

## Comprehensive Recommendations {.larger}

### Data Science Approach

**1. Data Quality Improvements:** - Collect modern car data - Include hybrid/electric vehicles - Expand feature set (aerodynamics, tech)

**2. Modeling Enhancements:** - Consider non-linear relationships - Explore interaction terms - Implement robust regression

**3. Practical Implementation:** - Regular model updates - Confidence interval reporting - Extrapolation warnings

**4. Future Considerations:** - Alternative algorithms (RF, GBM) - Cross-validation strategies - Feature engineering

## Questions?

Thank you for your attention!
