---
title: "Project 1"
author: "Mandip Chaudhary (17)"
date: "2025-03-09"
output:
  html_document: 
    toc: true
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session 1 and 2

## (Simple) Data entry in R

```{r}
data <- c(1,2,3,4,5,6,7,8,9)
data
text <- c("a", "b", "c", "d", "e", "f", "g", "h", "i")
text

# text2 <- c(a, b, c, d, e, f, g, h, i)
 # throws error as without quotes a,b,c,etc are treated as variables but are not defined

data2 <- cbind(data, text)
data2
```

```{r}
# data2 is defined as a two uni-dimensional matrix arrays i.e. data and text
class(data2)
```

## Arrays and Matrices in R:

```{r}
M <- matrix(
  c(1:9),
  nrow = 3,
  ncol = 3,
  byrow = TRUE
)

print(M)

```

```{r}
V <- c(1:12)

# Multidimensional array
MDA <- array(V, dim = c(2,3,2))
print(MDA)
```

## Creating a simple "data.frame" in R

```{r}
df <- data.frame(x=c(1,2,3), y=c(2,3,4), z=c(3,4,5))
print(df)
```

```{r}
print(class(df))
```

## A small but realistic data frame and its use:

```{r}
emp.data <- data.frame(
  emp_id = c(1:5),
  emp_name =c("Rick", "Dan", "Michelle", "Ryan", "Gary"),
  salary = c(623.3, 515.2, 611.0, 729.0, 845.25),
  start_date = as.Date(c("2012-01-01", "2013-09-23", "2014-11-15", "2014-05-
  11", "2015-03-27")),
  stringAsFactors = FALSE
)

#Print the data
print(emp.data)
```

## Structure and Summary of Sample Data Frame in R:

```{r}
str(emp.data)
```

### Print the Summary

```{r}
print(summary(emp.data))
```

## Extract part of data from Data Frame in R: (Very useful when working with large data)

```{r}
result <- data.frame(emp.data$emp_name, emp.data$salary)
print(result)
```

### Extract first two rows

```{r}
result <- emp.data[1:2,]
print(result)
```

### Extract 3rd and 5th row with 2nd and 4th column.

```{r}
result <- emp.data[c(3,5), c(2,4)]
print(result)
```

## Add a new column in existing Data Frame:

```{r}
# Add the "dept" column
emp.data$dept <- c("IT", "Operations", "IT", "HR", "Finance")
v <- emp.data
print(v)
```

## Expand data frame in R (Adding cases):

```{r}
# Create the first data frame.
emp.data <- data.frame(
  emp_id = c(1:5),
  emp_name = c("Rick", "Dan", "Michelle", "Ryan", "Gary"),
  salary = c(623.3, 515.2,611.0, 729.0,843.25),
  start_date = as.Date(c("2012-01-01", "2013-09-23", "2014-11-15", "2014-05-11", "2015-03-27")),
  dept = c("IT", "Operations", "IT", "HR", "Finance"),
  stringsAsFactors = FALSE
)

# Create the second data frame
emp.newdata <- data.frame(
  emp_id = c(6:8),
  emp_name = c("Ramsi", "Pranab", "Tusar"),
  salary = c(578.0,722.5,632.8),
  start_date = as.Date(c("2013-05-21", "2013-07-30", "2014-06-17")),
  dept = c("IT", "Operations","Finance"),
  stringsAsFactors = FALSE
)
emp.data
emp.newdata
```

## Expand data frame in R (rbind is used):

```{r}
# Bind the two data frames.
emp.finaldata <- rbind(emp.data, emp.newdata)
print(emp.finaldata)
```

## Import data in R: Text files

### Base read.csv function to read csv file:

```{r}
covid.data <- read.csv("~/Desktop/projects/r-starter/covnep_252days.csv")
head(covid.data)
```

## Import data in R: Excel files

-   Packages:
    -   "readxl", "xlxs" packages

### Readxl package to read excel files:

-   Get summary of this "data.frame" in R!

```{r}
library(readxl)
read_excel("~/Documents/Basic.xlsx")
```

# Session 3

## Basics of R

> R can do Math!. It follows PEMDAS rule. Parenthesis, Exponents, Multiplication, Division, Addition and Subtraction

```{r}
4 * 6 + 5 # 
(4 * 6) + 5 # 
4 * (6 + 5) #
(4 + 6)^2 * 5 / 10 + 9 - 1
```

## Variables in R: assigning and removing

-   Variable names can contain any combination of alphanumeric characters along with period(.) and underscore (\_) e.g. age.group or age_group
-   However, they cannot start with a number or an underscore e.g. \_age or 5age
-   Best practice is to use actual names, usually nouns for variables instead of single letter e.g. age, sex

```{r}
x <- 2 # (preferred)
x = 2
2 -> x
assign("x", 2)
rm(x)
```

## R is case sensitive

```{r}
theVariable <- 17
# will give error if we type: TheVariable or THE VARIABLE
Age <- 50
# will be different for: age or AGE
```

## Data Types

#### Numeric

-   Type of data can be checked using class() function
-   For numeric “class” and “is.numeric” both works:

```{r}
x <- c(1,2,3,4,5,6,7,8,9)
class(x)
is.numeric(x)
```

### Integer

-   For integer “class” and “is.numeric” both works:

```{r}
x <- c(1:9) # or c(1L:9L)
class(x)
X <- c(1L,2L,3L,4L,5L,6L,7L,8L,9L)
is.numeric(x) # R promotes “integers” to “numeric” when needed
```

#### R promotes “integers” to “numeric” when needed

```{r}
#Multiply integer by numeric in decimal values
4L * 2.8
class(4L)
class(2.8)
class(4L * 2.8)

# Divide integer by integer giving decimal value
5L / 2L # Will not promote to numeric here
class(5L)
class(2L)
class(5L / 2L)
4L * 5L  #Will also not promote here
2L + 4L + 5L

class(4L * 5L)
class(2L + 4L + 5L)
```

#### Character

```{r}
x <- "data"
x
class(x)
nchar(x) # number of characters


# Factor
y <- factor("data")
y
class(y)
# nchar(y) # Error in nchar(y) : 'nchar()' requires a character vectors
```

### Factors and attributes in R:

```{r}
gender <- factor(c("male", "female", "female", "male"))
typeof(gender) #datatype
attributes(gender) #Levels and class
unclass(gender) #Check how it is stored in R
```

### Date

```{r}
# Date: to store date
date1 <- as.Date("2023-03-29")
date1
class(date1)
as.numeric(date1)

# POSIXct: To store date and time
# Easier manipulation of date and time objects can be accomplished using “lubridate” and “chron” packages
date2 <- as.POSIXct("2023-03-29 06:30")
date2
class(date2)
as.numeric(date2)
```

#### Logical

```{r}
# TRUE (=1)
# FALSE (=0)

TRUE * 5
FALSE * 5

#Class and Check
k <- TRUE
class(k)
is.logical(k)
```

#### Logical Data Types

```{r}
2 == 3 # (FALSE)
2 != 3 # (TRUE)
2 < 3 # (TRUE)
2 <= 3 # (TRUE)
2 > 3 # (FALSE)
2 >= 3 # (FALSE)
"data" == "stats" # (FALSE, why?)
"data" < "stats" # (TRUE, why?)
```

#### Vectors

> A vector is collection of elements, all of the same type.

#### Vectors and its operation in R

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10) # x is a vector containing 10 elements
x <- c(1:10) # shortcut for above code
# c stands for “combine”
x * 3 #Multiplication by a scalar
x + 2 #Addition with a scalar
x - 3 #Subtraction with a scalar
x / 4 # Division by a scalar
x^2 #Exponentiation by a scalar
sqrt(x) #Square root
```

#### Extending vector operations in R

```{r}
#Two vector of equal length
x <- 1:10
y <- -5:4
x+y
x-y
x*y
x/y
x^y

#Check length of the vector
length(x)
length(y)
length(x+y)
```

### Extending vector operations in R

```{r}
#Two vectors of unequal length
x <- 1:10
z <-c(1,2)
x+z

#Two vectors of unequal length
x <- 1:10
w <-c(1,2,3)
x+w

#Comparing vectors
x <= 5
x > y
x < y

# Using “any” and “all”
x <- 10:1
y <- -4:5
any(x<y)
all(x<y)

#Using “nchar”
nchar(y)
```

### Assessing individual elements of a vector

```{r}
x[1]
# x[1,2] # throws error
x[c(1,2)] # correct syntax to get multiple elements of vector

#Giving names to a vector
#Name value pair
c(One="a", Two="y", Last="r")

#Create vector then name it
w <- 1:3
names(w) <- c("a", "b", "c")
w
```

## Missing data in R

-   R has two types of missing data
    -   NA
    -   NULL

### NA type missing data in R

```{r}
zchar <- c("Hockey", NA,"Cricket")
nchar(z)
z <- c(1,2,NA,8,3,NA,3)
mean(z)

#The “is.na” function tests each element of vector for missingness
is.na(z)

#The na.rm function with =TRUE argument will remove NA so that we can get values for:
mean(z, na.rm=TRUE)
var(z, na.rm=TRUE)
sd(z, na.rm=TRUE)
```

### NULL type missing data in R

```{r}
z <- c(1, NULL, 3) # NULL is the absence of anything
is.null(z)
d <- NULL
is.null(d)
is.null(7)
```

## Pipes in R

```{r}
x <- 1:10
mean(x)

library(magrittr)
# Mean of x with pipe:
x %>% mean
```

## Chained pipes in R

```{r}
z <-c(1,2,NA,8,3,NA,3)
sum(is.na(z))
#Pipes, without nesting
z %>% is.na %>% sum
#Additional argument
z %>% mean(na.rm=TRUE)
```

## Advanced data structures in R

```{r}
x <- 10:1
y <- -4:5
q <-c("Hockey", "Football","Baseball", "Kabaddi", "Rugby", "Pingpong", "Basketball","Tennis","Cricket", "Volleyball")
theDF <-data.frame(x, y, q)
theDF
```

## Advanced data structures in R

```{r}
theDF <-data.frame(First=x, Second=y, Sport=q)
names(theDF)
names(theDF)[3]
rownames(theDF)
rownames(theDF) <- c("One", "Two", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nice", "Ten")
rownames(theDF)
# Setting them back to generic index
rownames(theDF) <- NULL
rownames(theDF)

head(theDF)
head(theDF, n=7)
tail(theDF)
class(theDF)
str(theDF)
theDF[3,2]; theDF[3, 2:3]
theDF[, 3]; theDF[3,]
theDF[, c("First", "Sport")]
theDF[, "Sport", drop=FALSE]
```

## Lists in R

> Often a container is needed to hold arbitrary objects of either the same type or varying types. R accomplishes this through lists.

```{r}
list1 <- list(1,2,3)
list2 <- list(c(1,2,3))
list3 <- list(c(1,2,3), 3:7)
list4 <- list(theDF, 1:10)
list5 <- list(theDF, 1:10, list3)
names(list5)
names(list5) <-c("data.frame","vector", "list")
names(list5)
list5
list6 <- list(TheDataFrame=theDF,TheVector=1:10, TheList=list3)
names(list6)
```

### Access elements of list

```{r}
# Specify either the element number or name
list5[[1]]
list5[["data.frame"]]

#Accessed element manipulation
list5[[1]]$Sport #Sport variable
list5[[1]][, "Second"]
list5[[1]][, "Second", drop=F]
length(list5)
#Adding new element
list5[[4]] <- 2
list5[["NewElement"]] <-3:6
names(list5) ; list5;
```

## Matrices in R

-   This is a similar to a data.frame
-   It is rectangular with rows and columns except that every single element must be the same type, most commonly all numerics

```{r}
A <- matrix(1:10, nrow=5)
B <- matrix(21:30, nrow=5)
C <- matrix(21:40, nrow=2)
nrow(A)
ncol(B)
dim(C)
A + B; A * B; A - B; A = B;
```

## Matrix multiplication and names in R

```{r}
# Number of columns of the left hand matrix to be same as number of rows of right hand matrix
A %*% C # will work
# A %*% B # will not work

# Both A and B are 5 x 2 matrices so we will transpose B
A %*% t(B)

# Column/row names of matrix:
colnames(A)
colnames(A) <- c("Left", "Right")
rownames(A) <- c("1st", "2nd", "3rd", "4th", "5th")
t(A)
colnames(B) <- c("First","Second")
rownames(B) <- c("One", "Two", "Three", "Four", "Five")
```

## Arrays in R

-   An array is essentially a multidimensional vector

```{r}
theArray <- array(1:12,
dim=c(2,3,2)) # 2 dimensional matrices both with 2 rows and 3 columns

theArray [1, , ] # 1st row of both
theArray[1, ,1] # 1st row of first
theArray[,1,] # 1st column of both
```

## WEB SCRAPING WITH R

### Load in the Iris data from internet:

```{r}
iris <- read.csv(url("http://archive.ics.uci.edu/ml/machine-learning-databases/iris/iris.data"), header = FALSE)
head(iris)
```

#### Add column names for V1, V2, V3, V4 and V5 columns to the Iris data

```{r}
names(iris) <- c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width", "Species")
iris
```

## Saving the data frame as “csv” file in laptop:

```{r}
write.csv(iris, "iris.csv") #Will save CSV file in working directory
```

### store the current directory

```{r}
initial.dir <- getwd()
# change to the new Test directory
setwd("~/Desktop/projects/r-starter")

# load the necessary libraries
library(magrittr) #for pipes

# set the output file (it will bypass R and R Studio)
sink("session3.out")
# load the dataset from Test folder
iris <- read.csv("iris.csv")

# Do the analysis
plot(iris)

summary(iris)
iris %$% cor(Sepal.Length, Sepal.Width)
# close the output file
sink()

# unload the libraries
detach("package:magrittr")

# change back to the original directory
setwd(initial.dir)
```

## Using forward pipe operator/s in R:

> library(magrittr) required!

```{r}
library(magrittr)

# Compute the square root of `iris$Sepal.Length` and assign it to the new variable
iris$Sepal.Length.SQRT <- iris$Sepal.Length %>% sqrt()

# Compute the square root of `iris$Sepal.Length` and assign it to the same variable
iris$Sepal.Length %<>% sqrt

# Return `Sepal.Length`
iris$Sepal.Length
```

## The “tee” pipe operator “%T%”:

> library(magrittr) required!

```{r}
set.seed(123)
rnorm(200) %>%
matrix(ncol = 2) %T>%
plot %>%
colSums
```

## The exposing pipe operator “%\$%”:

> library(magrittr) required!

```{r}
iris %>%
subset(Sepal.Length > mean(Sepal.Length)) %$%
cor(Sepal.Length, Sepal.Width)

# The %$% operation comes handy for functions where “data” argument is not required/used like built-in “cor” function of R!
cor(iris$Sepal.Length, iris$Sepal.Width)
```

## Functions in R: Built-in functions

```{r}
# round()
round(3.1415)
round(3.1415, digits = 2)

# factorial()
factorial(3)
factorial(2*3)

# mean()
mean(1:6)
mean(c(1:30))
```

## “Sample” function: Random sampling without or with replacement in R

```{r}
die <- 1:6
sample(x = die, size = 1)
sample(x = die, size = 1)
sample (x = die, size = 1, replace=TRUE)
sample(x = die, size = 2)
sample(x = die, size = 2)
sample(x = die, size = 2, replace=TRUE)
```

## “Sample” function to split a datafile into train and test datasets

```{r}
# Make sure to have “iris.csv” datafile in the working directory and use read.csv to import it in R Studio:
read.csv("iris.csv")

# We can do the 70:30 random split of iris data frame as follow:
set.seed(123)
tt.sample <- sample(c(TRUE, FALSE), nrow(iris), replace=T, prob=c(0.7,0.3))
train <- iris[tt.sample, ]
test <- iris[!tt.sample, ]
```

## User-defined function in R:

```{r}
my_function <- function() {}
```

### User-defined function 1: roll()

```{r}
roll <- function() {
  die <- 1:6
  dice <- sample(die, size = 2, replace = TRUE)
  sum(dice)
}
roll()
roll()
roll()
```

### User-defined function 2: roll2()

```{r}
roll2 <- function(dice = 1:6) {
  dice <- sample(dice, size = 2, replace = TRUE)
  sum(dice)
}
roll2()
roll2()
roll2()
```

### User-defined function 3: roll3(data?)

```{r}
roll3 <- function(dice) {
  dice <- sample(dice, size = 2, replace = TRUE)
  sum(dice)
}

roll3(1:6)
roll3(1:12)
roll3(1:24)
```

### More Functions

```{r}
best_practice <- c("Let", "the", "computer", "do", "the", "work")
print_words <- function(sentence) {
  print(sentence[1])
  print(sentence[2])
  print(sentence[3])
  print(sentence[4])
  print(sentence[5])
  print(sentence[6])
}
print_words(best_practice) 
print_words(best_practice[-6]) 
best_practice[-6] 
```

## We can use functions with “for” loop in R!

```{r}
print_words <- function(sentence) {
  for (word in sentence) {
    print(word)
  }
}
print_words(best_practice) 
print_words(best_practice[-6]) 
best_practice[-6] 
```

## Condition: if and else

```{r}
#Checking values of y with x:
# if (y < 20) {
#   x <- "Too low"
# } else {
#  x <- "Too high"
# }

#Will this work?
check.y <- function(y) {
  if (y < 20) {
    print("Too Low")
  } else {
    print("Two high")
  }
}

check.y(10)
check.y(30)
```

## Creating binary variables with “ifelse”

```{r}
y <- 1:40
ifelse(y<20, "Too low", "Too high")
# It’s a logical as:
ifelse(y<20, TRUE, FALSE)

y <- 1:40
ifelse(y<20, 1, 0)
```

## Multiple conditions:

``` r
if (this) {
# do that
} else if (that) {
# do something else
} else if (that) {
# do something else
} else
# remaining
}
```

```{r}
check.x <- function(x=1:99){
  if (x<20){
    print("Less than 20")} else{
      if (x<40) {
        print("20-39")
      } else {
        if (x<100) {
          print("41-99")
        }
     }
   }
}
check.x(15)
check.x(30)
check.x(45)
```

## Multiple Conditions: combining “ifelse”

```{r}
x <- 1:99
x1 <- ifelse(x<20, 1,0) #Binary numbers
x2.1 <- ifelse(x<20, "<20", "20+") #Binary text
# x2.2 # ? For x between 20 and less than 40
# x2.3 # ? For x between 40 and less than 100

x3 <- ifelse(x<20,1,ifelse(x<40,2,3))
x3
table(x3)

#This code shows how Petal. Length categories was created from Petal. Length variable of iris data frame
iris <- within(iris, {
  Petal.cat <- NA
  Petal.cat[Petal.Length <1.6] <- "Small"
  Petal.cat[Petal.Length >=1.6 &
  Petal.Length<5.1] <- "Medium"
  Petal.cat[Petal.Length >=5.1] <- "Large"
})
#The 1.6=Q1 and 5.1=Q3 were obtained from the “summary” of the Petal.Lenght variable i.e. summary(iris$Petal.Length)
iris$Petal.cat
table(iris$Petal.cat)
```

## Multiple Conditions: If, else if, else if, else if

``` r
if (temp <= 0) {
"freezing"}
else if (temp <= 10) {
"cold"}
else if (temp <= 20) {
"cool"}
else if (temp <= 30) {
"warm"}
else {
"hot"}
```

```{r}
temp_description <- function(temp) {
  if (temp <= 0) {
    "freezing"
  } else if (temp <= 10) {
    "cold"
  } else if (temp <= 20) {
    "cool"
  } else if (temp <= 30) {
    "warm"
  } else {
    "hot"
  }
}

# Example usage:
temp_description(0)
temp_description(10)
temp_description(15)
temp_description(25)
temp_description(35)
```

## Scraping Covid data from Wikipedia and html parsing

```{r}
library(rvest)
library(dbplyr)

url <- "https://en.wikipedia.org/wiki/COVID-19_pandemic_in_Nepal#Data"

covid_data <- read_html(url)

table_data <- covid_data %>% html_nodes(".COVID-19_pandemic_data_Nepal_medical_cases table") %>% html_table()

head(table_data)

table_df <- as.data.frame(table_data)
table_df
names(table_df)

table_df$Date

head(table_data)

names(table_df)

table.tibble <- tibble::as_tibble(table_df)
```
### DUCK DB
```{r}
library(duckdb)
con <- DBI::dbConnect()
```



