---
title: "Decision Tree Classifier"
output: html_notebook
---

```{r}
# ignore warning
options(warn=-1)
```


```{r}
library(readr)
```


```{r}
data <- read_csv("~/Desktop/r-starter/Cardiotocographic.csv")
str(data)
```


```{r}
data$NSPF <- as.factor(data$NSP)
str(data$NSPF)
```


```{r}
set.seed(17)
library(caret)
library(party)

ind = sample(2, nrow(data), replace = TRUE, prob = c(0.7, 0.3))
train = data[ind == 1, ]
test = data[ind == 2, ]

# Fit the decision tree model with only the first three varibale to learn it today 
tree <- ctree(NSPF ~ LB+AC+FM, data=train)
tree
```

```{r}
plot(tree)
```

## Pruning the tree with 99% confidence interval and split at 500 samples
```{r}
tree1 <- ctree(NSPF ~ LB+AC+FM, data=train, controls = ctree_control(minsplit = 500, mincriterion = 0.99))
plot(tree1)
```

```{r}
pred <- predict(tree, type="prob")
```


```{r}
predict(tree, test)
```
# Confusion matrix
```{r}
(tab <- table(predict(tree), train$NSPF))
```
# Accuracy
```{r}
accuracy <- sum(diag(tab)) / sum(tab)
accuracy
```

# Misclassification error
```{r}
mce <- 1 - accuracy
mce
```

# Prediction for test data
```{r}
pred.test <- predict(tree, newdata=test)
```

# Confusion Matrix
```{r}
(tab1 <- table(pred.test, test$NSPF))
```
# Accuracy
```{r}
accuracy.test <- sum(diag(tab1)) / sum(tab1)
accuracy.test
```

# Misclassification error
```{r}
mce.test <- 1 - accuracy.test
mce.test
```

# Bagging in R with "ipred" package for CTG data (NSP = dependent variable)
```{r}
# Bagging with ipred package
library(ipred)
MBTree <- bagging(NSPF ~ LB+AC+FM, data=train, coob=TRUE)
MBTree
```
# Prediction
```{r}
MBPrediction <- predict(MBTree, test)
MBPrediction
```

# Confusion Matrix
```{r}
confusionMatrix(MBPrediction, test$NSPF)
```

# Random Forest
```{r}
library(randomForest)
set.seed(17)
# Fitting the random forest model
rfm <- randomForest(NSPF ~ LB+AC+FM, data=train)
rfm
```

```{r}
library(caret)
# Prediction
rfmPrediction <- predict(rfm, train)
head(rfmPrediction)
```

```{r}
head(train$NSPF)
```

# Confusion Matrix
```{r}
confusionMatrix(rfmPrediction, train$NSPF)
```

# Prediction for test data
```{r}
rfmPredictionTest <- predict(rfm, test)
# Confusion Matrix
confusionMatrix(rfmPredictionTest, test$NSPF)
```

```{r}
plot(rfm)
```

# We can use "tuneRF" function of the randomForest package but we need to tdefine train as data frame
```{r}
train <- as.data.frame(train)

t <- tuneRF(train[,-22], train[,22], ntree=200, ,mtry=4, importance = TRUE, proximity = TRUE, doBest = TRUE, plot=TRUE)
t
```
```{r}
test <- as.data.frame(test)

t <- tuneRF(test[,-22], test[,22], ntree=200, ,mtry=4, importance = TRUE, proximity = TRUE, doBest = TRUE, plot=TRUE)
t
```

# Generalized Boosting Regression Model
```{r}
library(caret)
mod.gbm <- train(NSPF ~ ., data=train, method="gbm", verbose=FALSE)
print(mod.gbm)
```

```{r}
library(caret)
mod.xgb <- train(NSPF ~ ., data=train, method="xgbTree", verbose=TRUE)
```

```{r}
print(mod.xgb)
summary(mod.xgb)
```


































