---
title: "R Notebook"
output: html_notebook
---
## Example 1: Test the claim that mean of the miles per gallon variable is 20 using "mtcars" data in R!
### Hypothesis
- Null hypothesis: H0: μ = 20

```{r}
mu0 <- 20
sigma <- 6
xbar <- mean(mtcars$mpg)
n <- length(mtcars$mpg)
```

```{r}
z <- sqrt(n) * (xbar - mu0) / sigma
p_value <- 2 * (1 - pnorm(abs(z)))

print(p_value)

p_value <- 2 * pnorm(z)

print(p_value)
```

## Saffrio test 
```{r}
shapiro.test(mtcars$mpg)
```

## T-test

```{r}
t.test(mtcars$mpg, mu = mu0)
```

```{r}
t.test(mtcars$mpg, mu = 10)
```

```{r}
t.test(mtcars$mpg, mu = 30)
```


```{r}
?mtcars
```

## Normality test on am variable of mtcars dataset
```{r}
var.test(mpg ~ am,data=mtcars)
```

```{r}
with(mtcars, shapiro.test(mpg[am==0]))
```
```{r}
with(mtcars, shapiro.test(mpg[am==1]))
```

```{r}
var.test(mpg~am, data=mtcars)
```
```{r}
t.test(mpg~am, var.equal=T, data=mtcars)
```

```{r}
summary(lm(mpg~am, data=mtcars))
```

```{r}
plot(mtcars$am, mtcars$mpg)
```

```{r}
cor.test(mtcars$mpg, mtcars$am)
```

Linear regression in extension on t-test.
Logistic regression is an extension of Chi-square test

CHI Square test Assumptions
- The sample size of the dataset must be greater than 50.
- The continuity correction is done for normal approximation for binomial distribution.

```{r}
table(mtcars$gear)
```


## 1-way ANOVA assumptions checks:

### 1. Normality by categories

```{r}
with(mtcars, shapiro.test(mpg[gear==3]))
```
```{r}
with(mtcars, shapiro.test(mpg[gear==4]))
```
```{r}
with(mtcars, shapiro.test(mpg[gear==5]))
```

### 2. Equal variabce among categories:
```{r}
library(car)
leveneTest(mpg~factor(gear), data=mtcars)
```
## Multiple proportion test

```{r}
table(mtcars$gear)
```

```{r}
prop.test(x=c(15,12,5), n=c(32, 32, 32))
```
```{r}
prop.test(x=c(15,12,5), n=c(32, 32, 32), correct = F)
```
```{r}
pairwise.prop.test(x=c(15,12,5), n=c(32, 32, 32), correct = F)
```

```{r}

```












































