---
title: "Decision Tree and Random Forest Models"
author: "Roll Number: 17"
date: "`r Sys.Date()`"
output: html_notebook
---

## Data Generation and Preparation

First, we'll generate synthetic data with 500 observations and 5 variables. One variable will be a categorical factor with levels 1, 2, and 3 as the dependent variable.
a) code used to generate 500 random data with five variables where one of the variables must be defined as categorical factor (1, 2 and 3) dependent variable

```{r setup, message=FALSE}
# Load required packages
library(rpart)        # For decision trees
library(randomForest) # For random forest
library(caret)        # For model evaluation
library(ggplot2)      # For visualization
```

b) code used to fix the random data with random seed i.e. your roll number

```{r data_generation}
# Set random seed using roll number
set.seed(17)

# Generate 4 continuous predictor variables
x1 <- rnorm(500, mean = 0, sd = 1)  # Normal distribution with mean 0, sd 1
x2 <- rnorm(500, mean = 2, sd = 2)  # Normal distribution with mean 2, sd 2
x3 <- rnorm(500, mean = -1, sd = 1.5)  # Normal distribution with mean -1, sd 1.5
x4 <- rnorm(500, mean = 1, sd = 1)  # Normal distribution with mean 1, sd 1

# Generate response variable as a function of predictors with little randomness
# This approach is used to ensure that the response variable is mostly determined by the predictors,
# making it easier for machine learning models to learn the underlying pattern and achieve higher accuracy.
# Adding a small amount of noise (randomness) makes the data more realistic, but the main relationship remains strong.
y <- ifelse(x1 + x2 + rnorm(500, 0, 0.5) > 2, 1,
     ifelse(x3 - x4 + rnorm(500, 0, 0.5) < 0, 2, 3))
y <- factor(y, levels = c(1, 2, 3))
# Create data frame
data <- data.frame(
  x1 = x1,
  x2 = x2,
  x3 = x3,
  x4 = x4,
  result = y
)

# Split data into training (70%) and testing (30%) sets
set.seed(17)  # For reproducibility of split
ind<-sample(2, nrow(data), prob=c(0.7,0.3), replace = T)
train_data <- data[ind==1, ]
test_data <- data[ind==2, ]
```

**Description**: We generated 500 observations with 4 continuous predictor variables (x1, x2, x3, x4) from different normal distributions and 1 categorical response variable with three levels (1, 2, 3). The data was split into training (70%) and testing (30%) sets.

c) code used to fit the assigned supervised classification and unsupervised statistical models 

## Decision Tree Model

```{r decision_tree}
par(mar = c(5.1, 4.1, 6.1, 2.1)) # Increased top margin to 6.1

# Fit decision tree model
tree_model <- rpart(result ~ ., data = train_data, method = "class")

# Plot decision tree
plot(tree_model, uniform = TRUE, main = "Decision Tree")
text(tree_model, use.n = TRUE, all = TRUE, cex = 0.8)
```

```{r}
# Make predictions on test set
tree_pred <- predict(tree_model, test_data, type = "class")
# Calculate confusion matrix and accuracy metrics
tree_cm <- confusionMatrix(tree_pred, test_data$result)

# Print results
print("Decision Tree Model Results:")
print(tree_cm)
```

**Interpretation**: The decision tree model shows how the predictor variables are used to classify the response variable into three categories. The confusion matrix provides insights into model performance, showing how well the model predicted each class. The accuracy metric indicates the overall correctness of predictions.

## Random Forest Model

```{r random_forest}
# Fit random forest model
rf_model <- randomForest(result ~ ., data = train_data, ntree = 500)

# Make predictions on test set
rf_pred <- predict(rf_model, test_data)

# Calculate confusion matrix and accuracy metrics
rf_cm <- confusionMatrix(rf_pred, test_data$result)

# Print results
print("Random Forest Model Results:")
print(rf_cm)
```

```{r}
# Plot variable importance
varImpPlot(rf_model, main = "Variable Importance Plot")
```

**Interpretation**: The random forest model, being an ensemble method, typically provides more robust predictions than a single decision tree. The variable importance plot shows which predictors are most influential in making predictions. The confusion matrix and accuracy metrics can be compared with the decision tree model to evaluate which performs better.

## Model Comparison and Conclusions

```{r model_comparison}
# Create comparison table
comparison <- data.frame(
  Model = c("Decision Tree", "Random Forest"),
  Accuracy = c(tree_cm$overall["Accuracy"], rf_cm$overall["Accuracy"]),
  Kappa = c(tree_cm$overall["Kappa"], rf_cm$overall["Kappa"])
)

# Print comparison
print(comparison)
```

**Note**:
Kappa (Cohen's Kappa) is a statistical measure of inter-rater agreement or classification accuracy that accounts for agreement occurring by chance. In the context of classification models, Kappa compares the observed accuracy with the expected accuracy (random chance).
- Kappa = 1: Perfect agreement between predictions and actual values.
- Kappa = 0: Agreement is no better than random chance.
- Kappa < 0: Agreement is worse than chance.

**Interpretation**:

1. **Model Accuracy**: The comparison shows which model achieved higher accuracy on the test set. Random Forest typically performs better due to its ensemble nature.

2. **Model Complexity**: 
   - Decision Tree: Simpler, more interpretable model with clear decision rules
   - Random Forest: More complex but typically more accurate, with built-in feature importance

3. **Practical Implications**:
   - If interpretability is crucial, the decision tree might be preferred
   - If prediction accuracy is the primary goal, random forest would be the better choice

4. **Variable Importance**: The random forest's variable importance plot helps identify which predictors are most useful for classification, which can guide feature selection in future modeling efforts.
