---
title: "Decision Tree and Random Forest Models with Student Performance Data"
author: "Roll Number: 17"
date: "`r Sys.Date()`"
output: html_notebook
---

## Data Generation and Preparation

We'll generate synthetic student performance data with 500 observations.

```{r setup, message=FALSE}
# Load required packages
library(rpart)        # For decision trees
library(rpart.plot)   # For better tree visualization
library(randomForest) # For random forest
library(caret)        # For model evaluation and tuning
library(pROC)        # For ROC curves
```

```{r data_generation}
# Set random seed using roll number
set.seed(17)

# Generate 4 continuous predictor variables
study_hours <- rnorm(500, mean=5, sd=1.5)     # Study hours per day
prev_scores <- rnorm(500, mean=75, sd=10)     # Previous scores
attendance <- rnorm(500, mean=85, sd=8)        # Attendance percentage
parent_edu <- rnorm(500, mean=14, sd=2)       # Parent's education years

# Generate performance level based on predictors
# Create a score that combines all predictors with some randomness
performance_score <- 0.3*scale(study_hours) + 
                    0.3*scale(prev_scores) + 
                    0.2*scale(attendance) + 
                    0.2*scale(parent_edu) + 
                    rnorm(500, 0, 0.5)

# Create data frame
data <- data.frame(
  study_hours = study_hours,
  prev_scores = prev_scores,
  attendance = attendance,
  parent_edu = parent_edu,
  performance = performance_level
)

# Split data into training (70%) and testing (30%) sets
set.seed(17)
train_index <- createDataPartition(data$performance, p = 0.7, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]
```

**Description**: Generated student performance dataset with 4 predictors and a categorical response variable (performance levels 1,2,3). The predictors influence the performance level with some added randomness to make it realistic.

## Base Decision Tree Model

```{r base_decision_tree}
# Fit initial decision tree model
tree_model <- rpart(performance ~ ., 
                   data = train_data, 
                   method = "class",
                   control = rpart.control(cp = 0.01))
# Plot the tree
rpart.plot(tree_model, 
           main = "Initial Decision Tree",
           extra = 1,
           box.palette = "skyblue")

# Find optimal cp
opt_cp <- tree_model$cptable[which.min(tree_model$cptable[,"xerror"]),"CP"]
print(paste("Optimal complexity parameter:", opt_cp))

# Prune tree
pruned_tree <- prune(tree_model, cp = opt_cp)

rpart.plot(pruned_tree, 
           main = "Initial Decision Tree",
           extra = 1,
           box.palette = "skyblue")

# Make predictions
tree_pred <- predict(tree_model, test_data, type = "class")

# Calculate performance metrics
tree_cm <- confusionMatrix(tree_pred, test_data$performance)

# Print results
print("Base Decision Tree Results:")
print(tree_cm)
```

## Tree Pruning

```{r tree_pruning}
# Cross-validation for cp selection
printcp(tree_model)
plotcp(tree_model)

# Find optimal cp
opt_cp <- tree_model$cptable[which.min(tree_model$cptable[,"xerror"]),"CP"]
print(paste("Optimal complexity parameter:", opt_cp))

# Prune tree
pruned_tree <- prune(tree_model, cp = opt_cp)

# Plot pruned tree
rpart.plot(pruned_tree, 
           main = "Pruned Decision Tree",
           extra = 1,
           box.palette = "skyblue")

# Make predictions with pruned tree
pruned_pred <- predict(pruned_tree, test_data, type = "class")

# Calculate performance metrics
pruned_cm <- confusionMatrix(pruned_pred, test_data$performance)

# Print results
print("Pruned Decision Tree Results:")
print(pruned_cm)
```

## Random Forest Model

```{r random_forest}
# Initial random forest model
rf_model <- randomForest(performance ~ ., 
                        data = train_data,
                        ntree = 500,
                        importance = TRUE)

# Make predictions
rf_pred <- predict(rf_model, test_data)

# Calculate performance metrics
rf_cm <- confusionMatrix(rf_pred, test_data$performance)

# Print results
print("Initial Random Forest Results:")
print(rf_cm)

# Plot variable importance
varImpPlot(rf_model, main = "Variable Importance Plot")
```

## Random Forest Hyperparameter Tuning

```{r rf_tuning}
# Define parameter grid
tuning_grid <- expand.grid(
  mtry = 1:4,  # number of variables at each split
  splitrule = "gini",
  min.node.size = c(1, 3, 5)
)

# Set up cross-validation
ctrl <- trainControl(method = "cv",
                    number = 5,
                    search = "grid")

# Train model with cross-validation
rf_tuned <- train(performance ~ .,
                  data = train_data,
                  method = "ranger",
                  trControl = ctrl,
                  tuneGrid = tuning_grid,
                  importance = "impurity")

# Print best parameters
print("Best Tuning Parameters:")
print(rf_tuned$bestTune)

# Make predictions with tuned model
tuned_pred <- predict(rf_tuned, test_data)

# Calculate performance metrics
tuned_cm <- confusionMatrix(tuned_pred, test_data$performance)

# Print results
print("Tuned Random Forest Results:")
print(tuned_cm)
```

## Model Comparison

```{r model_comparison}
# Create comparison table
comparison <- data.frame(
  Model = c("Base Decision Tree", "Pruned Decision Tree", 
            "Initial Random Forest", "Tuned Random Forest"),
  Accuracy = c(tree_cm$overall["Accuracy"],
              pruned_cm$overall["Accuracy"],
              rf_cm$overall["Accuracy"],
              tuned_cm$overall["Accuracy"]),
  Kappa = c(tree_cm$overall["Kappa"],
            pruned_cm$overall["Kappa"],
            rf_cm$overall["Kappa"],
            tuned_cm$overall["Kappa"])
)

# Print comparison
print("Model Comparison:")
print(comparison)

# Plot ROC curves (one-vs-all for each class)
par(mfrow=c(2,2))
```

**Final Interpretation**:

1. **Model Evolution**:
   - Base Decision Tree: Shows initial classification structure
   - Pruned Tree: Reduces overfitting while maintaining performance
   - Random Forest: Improves accuracy through ensemble learning
   - Tuned Random Forest: Further optimizes model parameters

2. **Performance Comparison**:
   - Accuracy progression across models
   - Sensitivity and specificity for each class
   - ROC curves showing model discrimination ability

3. **Variable Importance**:
   - Study hours and previous scores typically most influential
   - Attendance and parent education provide additional predictive power

4. **Practical Implications**:
   - Decision trees offer interpretability
   - Random forest provides higher accuracy
   - Pruning and tuning demonstrate model optimization
   - Trade-off between model complexity and performance
