---
title: "Decision Tree and Random Forest Models"
author: "Roll Number: 17"
date: "`r Sys.Date()`"
output: html_notebook
---

## Data Generation and Preparation

First, we'll generate synthetic data with 500 observations and 5 variables. One variable will be a categorical factor with levels 1, 2, and 3 as the dependent variable.
a) code used to generate 500 random data with five variables where one of the variables must be defined as categorical factor (1, 2 and 3) dependent variable

```{r setup, message=FALSE}
# Load required packages
library(rpart)        # For decision trees
library(rpart.plot)   # For tree visualization
library(randomForest) # For random forest
library(caret)        # For model evaluation and tuning
library(ggplot2)      # For visualization
library(pROC)        # For ROC curves and AUC
```

b) code used to fix the random data with random seed i.e. your roll number

```{r data_generation}
# Set random seed using roll number
set.seed(17)

# Generate 4 continuous predictor variables
x1 <- rnorm(500, mean = 0, sd = 1)  # Normal distribution with mean 0, sd 1
x2 <- rnorm(500, mean = 2, sd = 2)  # Normal distribution with mean 2, sd 2
x3 <- rnorm(500, mean = -1, sd = 1.5)  # Normal distribution with mean -1, sd 1.5
x4 <- rnorm(500, mean = 1, sd = 1)  # Normal distribution with mean 1, sd 1
# Generate response variable as a function of predictors with little randomness
# This approach is used to ensure that the response variable is mostly determined by the predictors,
# making it easier for machine learning models to learn the underlying pattern and achieve higher accuracy.
# Adding a small amount of noise (randomness) makes the data more realistic, but the main relationship remains strong.
y <- ifelse(x1 + x2 + rnorm(500, 0, 0.5) > 2, 1,
     ifelse(x3 - x4 + rnorm(500, 0, 0.5) < 0, 2, 3))
# Convert numeric levels to valid R names
y <- factor(y, levels = c(1, 2, 3), 
           labels = c("Low", "Medium", "High"))
# Create data frame
data <- data.frame(
  x1 = x1,
  x2 = x2,
  x3 = x3,
  x4 = x4,
  result = y
)

# Split data into training (70%) and testing (30%) sets
set.seed(17)  # For reproducibility of split
ind<-sample(2, nrow(data), prob=c(0.7,0.3), replace = T)
train_data <- data[ind==1, ]
test_data <- data[ind==2, ]
```

**Description**: We generated 500 observations with 4 continuous predictor variables (x1, x2, x3, x4) from different normal distributions and 1 categorical response variable with three levels (1, 2, 3). The data was split into training (70%) and testing (30%) sets.

c) code used to fit the assigned supervised classification and unsupervised statistical models 

## Decision Tree Model

```{r decision_tree}
# Fit initial decision tree model with lower complexity parameter
tree_model <- rpart(result ~ ., 
                   data = train_data, 
                   method = "class", cp=0.01)  # Even smaller cp for initial tree

# Plot initial tree
# par(mar = c(5.1, 4.1, 6.1, 2.1))
rpart.plot(tree_model, 
           main = "Initial Complex Tree")
```
```{r}
# Choose the CP value with lowest cross-validated error
best_cp <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"]
best_cp

# Prune the tree using the best CP value
pruned_tree <- prune(tree_model, cp = 0.01136364)

# Plot the pruned tree
rpart.plot(pruned_tree)

# Make predictions with pruned tree
pruned_pred <- predict(pruned_tree, test_data, type = "class")

# Calculate performance metrics
pruned_cm <- confusionMatrix(pruned_pred, test_data$result)

# Print results
print("Pruned Decision Tree Results:")
print(pruned_cm)

# Compare tree structures
cat("\nTree Complexity Comparison:\n")
cat("Initial tree nodes:", nrow(tree_model$frame), "\n")
cat("Pruned tree nodes:", nrow(pruned_tree$frame), "\n")
cat("Node reduction:", nrow(tree_model$frame) - nrow(pruned_tree$frame), "nodes\n")
cat("Complexity reduction:", 
    round((nrow(tree_model$frame) - nrow(pruned_tree$frame))/nrow(tree_model$frame) * 100, 1),
    "%\n")
```


```{r}
# Print initial tree complexity
cat("\nInitial Tree Size:\n")
cat("Number of nodes:", nrow(tree_model$frame), "\n")

# Make predictions
tree_pred <- predict(tree_model, test_data, type = "class")

# Calculate performance metrics
tree_cm <- confusionMatrix(tree_pred, test_data$result)

# Print results
print("Base Decision Tree Results:")
print(tree_cm$byClass)

# Cross-validation for cp selection
printcp(tree_model)
plotcp(tree_model)
```

**Interpretation**: The decision tree analysis shows:
1. Initial Tree: A complex tree with many nodes (cp = 0.0001) to capture fine details in the data
2. Pruned Tree: A simplified version that maintains performance while reducing overfitting
3. The pruning process significantly reduces tree complexity while preserving predictive power
4. The confusion matrix and accuracy metrics show the performance trade-offs between complexity and accuracy

The significant reduction in tree size through pruning demonstrates the importance of finding the right balance between model complexity and predictive accuracy.

## Random Forest Model

```{r random_forest}
# Initial random forest model
rf_model <- randomForest(result ~ ., 
                        data = train_data,
                        ntree = 500,
                        importance = TRUE)

# Make predictions
rf_pred <- predict(rf_model, test_data)

# Calculate performance metrics
rf_cm <- confusionMatrix(rf_pred, test_data$result)

# Print results
print("Initial Random Forest Results:")
print(rf_cm)

# Plot variable importance
varImpPlot(rf_model, 
           main = "Random Forest Variable Importance",
           type = 1,
           cex = 0.8)
```

## Random Forest Hyperparameter Tuning

```{r rf_tuning}
# Function to perform k-fold cross-validation
rf_cv <- function(data, k=5, mtry) {
    set.seed(17)
    folds <- sample(1:k, nrow(data), replace=TRUE)
    cv_errors <- numeric(k)
    
    for(i in 1:k) {
        # Split data
        train <- data[folds != i,]
        test <- data[folds == i,]
        
        # Train model
        rf <- randomForest(result ~ ., 
                          data=train, 
                          mtry=mtry,
                          ntree=500,
                          importance=TRUE)
        
        # Make predictions
        pred <- predict(rf, test)
        
        # Calculate error
        cv_errors[i] <- mean(pred != test$result)
    }
    
    return(mean(cv_errors))
}

# Try different mtry values
mtry_values <- 1:4
cv_results <- sapply(mtry_values, function(m) rf_cv(train_data, mtry=m))

# Find best mtry
best_mtry <- mtry_values[which.min(cv_results)]
print(paste("Best mtry value:", best_mtry))

# Train final model with best mtry
rf_tuned <- randomForest(result ~ ., 
                        data=train_data,
                        mtry=best_mtry,
                        ntree=500,
                        importance=TRUE)

# Make predictions
tuned_pred <- predict(rf_tuned, test_data)

# Calculate performance metrics
tuned_cm <- confusionMatrix(tuned_pred, test_data$result)

# Print results
print("Tuned Random Forest Results:")
print(tuned_cm)

# Plot error rates
plot(rf_tuned, main="Error Rates vs Number of Trees")

# Plot variable importance
varImpPlot(rf_tuned, 
           main="Variable Importance (Tuned Model)",
           type=1,
           cex=0.8)

# Print tuning results
cat("\nCross-validation results:\n")
results_df <- data.frame(
    mtry = mtry_values,
    cv_error = cv_results
)
print(results_df)
```

**Interpretation**: The random forest analysis demonstrates:
1. Initial Model: Built with 500 trees, providing robust ensemble predictions
2. Variable Importance: Shows relative contribution of each predictor to classification
3. Performance Comparison: Generally outperforms single decision trees due to ensemble nature
4. Cross-validation Results: Shows model stability across different data subsets

The random forest approach provides a more robust solution by combining multiple decision trees, though at the cost of reduced interpretability compared to single trees.

## Model Comparison and Conclusions

```{r model_comparison}
# Create comprehensive comparison table
comparison <- data.frame(
  Model = c("Base Decision Tree", "Pruned Decision Tree", 
            "Initial Random Forest", "Tuned Random Forest"),
  Accuracy = c(tree_cm$overall["Accuracy"],
              pruned_cm$overall["Accuracy"],
              rf_cm$overall["Accuracy"],
              tuned_cm$overall["Accuracy"]),
  Kappa = c(tree_cm$overall["Kappa"],
            pruned_cm$overall["Kappa"],
            rf_cm$overall["Kappa"],
            tuned_cm$overall["Kappa"])
)

# Print overall performance comparison
print("Overall Model Performance:")
print(comparison)
tree_cm
# Create class-wise performance comparison
performance_metrics <- data.frame(
    Class = rep(c("Low", "Medium", "High"), each = 4),
    Model = rep(c("Base Tree", "Pruned Tree", "Random Forest", "Tuned RF"), times = 3),
    Sensitivity = c(
        # Low class
        tree_cm$byClass["Low", "Sensitivity"],
        pruned_cm$byClass["Low", "Sensitivity"],
        rf_cm$byClass["Low", "Sensitivity"],
        tuned_cm$byClass["Low", "Sensitivity"],
        # Medium class
        tree_cm$byClass["Medium", "Sensitivity"],
        pruned_cm$byClass["Medium", "Sensitivity"],
        rf_cm$byClass["Medium", "Sensitivity"],
        tuned_cm$byClass["Medium", "Sensitivity"],
        # High class
        tree_cm$byClass["High", "Sensitivity"],
        pruned_cm$byClass["High", "Sensitivity"],
        rf_cm$byClass["High", "Sensitivity"],
        tuned_cm$byClass["High", "Sensitivity"]
    ),
    Specificity = c(
        # Low class
        tree_cm$byClass["Low", "Specificity"],
        pruned_cm$byClass["Low", "Specificity"],
        rf_cm$byClass["Low", "Specificity"],
        tuned_cm$byClass["Low", "Specificity"],
        # Medium class
        tree_cm$byClass["Medium", "Specificity"],
        pruned_cm$byClass["Medium", "Specificity"],
        rf_cm$byClass["Medium", "Specificity"],
        tuned_cm$byClass["Medium", "Specificity"],
        # High class
        tree_cm$byClass["High", "Specificity"],
        pruned_cm$byClass["High", "Specificity"],
        rf_cm$byClass["High", "Specificity"],
        tuned_cm$byClass["High", "Specificity"]
    )
)

# Round numeric columns to 3 decimal places
performance_metrics[,3:4] <- round(performance_metrics[,3:4], 3)

# Print class-wise performance comparison
cat("\nClass-wise Performance Metrics:\n")
print(knitr::kable(performance_metrics, 
                   caption = "Sensitivity and Specificity by Class and Model",
                   align = c('l', 'l', 'r', 'r')))

# Add interpretation
cat("\nInterpretation of class-wise performance:\n")
cat("- Sensitivity: Ability to correctly identify samples of that class\n")
cat("- Specificity: Ability to correctly identify samples not in that class\n")
cat("- Higher values indicate better performance (range: 0 to 1)\n\n")
cat("Key observations:\n")
cat("1. Class-wise Performance: Each class (Low, Medium, High) shows different patterns\n")
cat("2. Model Comparison: Shows how each model performs for specific classes\n")
cat("3. Trade-offs: Highlights balance between sensitivity and specificity\n")
cat("4. Model Selection: Helps choose models based on class-specific needs\n")

# Round numeric columns to 3 decimal places
comparison[,-1] <- round(comparison[,-1], 3)

# Print comparison with nice formatting
print("Model Performance Comparison:")
print(comparison)

# Plot ROC curves for each class
par(mfrow=c(2,2))

# For each class level
for(class_level in c("Low", "Medium", "High")) {
    # Create binary outcomes
    binary_actual <- ifelse(test_data$result == class_level, 1, 0)
    
    # Get probabilities for each model
    tree_probs <- predict(tree_model, test_data, type="prob")[,class_level]
    pruned_probs <- predict(pruned_tree, test_data, type="prob")[,class_level]
    rf_probs <- predict(rf_model, test_data, type="prob")[,class_level]
    rf_tuned_probs <- predict(rf_tuned, test_data, type="prob")[,class_level]
    
    # Create ROC curves
    roc_tree <- roc(binary_actual, tree_probs)
    roc_pruned <- roc(binary_actual, pruned_probs)
    roc_rf <- roc(binary_actual, rf_probs)
    roc_tuned <- roc(binary_actual, rf_tuned_probs)
    
    # Plot ROC curves
    plot(roc_tree, main=paste("ROC for", class_level, "Class"),
         col="blue", lwd=2)
    lines(roc_pruned, col="green", lwd=2)
    lines(roc_rf, col="red", lwd=2)
    lines(roc_tuned, col="purple", lwd=2)
    
    # Add legend
    legend("bottomright", 
           legend=c("Base Tree", "Pruned Tree", 
                   "Random Forest", "Tuned RF"),
           col=c("blue", "green", "red", "purple"), 
           lwd=2)
    
    # Add AUC values
    auc_text <- sprintf("AUC: %.3f / %.3f / %.3f / %.3f",
                       auc(roc_tree), auc(roc_pruned),
                       auc(roc_rf), auc(roc_tuned))
    text(0.5, 0.2, auc_text, cex=0.8)
}

# Reset plot layout
par(mfrow=c(1,1))

```

**Note**:
Kappa (Cohen's Kappa) is a statistical measure of inter-rater agreement or classification accuracy that accounts for agreement occurring by chance. In the context of classification models, Kappa compares the observed accuracy with the expected accuracy (random chance).
- Kappa = 1: Perfect agreement between predictions and actual values.
- Kappa = 0: Agreement is no better than random chance.
- Kappa < 0: Agreement is worse than chance.

**Final Analysis and Interpretation**:

1. **Model Evolution and Performance**:
   - Base Decision Tree: Provides a baseline model with clear decision rules
   - Pruned Decision Tree: Reduces complexity while maintaining performance, preventing overfitting
   - Initial Random Forest: Leverages ensemble learning for improved accuracy
   - Tuned Random Forest: Further optimizes performance through hyperparameter tuning

2. **Performance Metrics Comparison**:
   - Overall Metrics:
     * Accuracy: Measures overall correct predictions
     * Kappa: Agreement accounting for chance, showing model reliability
   - Class-specific Metrics:
     * Low Class: Performance in identifying and distinguishing low values
     * Medium Class: Performance in identifying and distinguishing medium values
     * High Class: Performance in identifying and distinguishing high values
   - Each class evaluated by:
     * Sensitivity: True positive rate for that class
     * Specificity: True negative rate for that class

3. **ROC Curve Analysis**:
   - Demonstrates model performance across different classification thresholds
   - Each class's separate ROC curve shows model's discrimination ability
   - Area Under Curve (AUC) quantifies overall model performance
   - Comparison across models shows relative strengths

4. **Hyperparameter Optimization**:
   - Tree Pruning: Optimal complexity parameter balances model complexity and accuracy
   - Random Forest Tuning: Optimized mtry and min.node.size improve model robustness
   - Cross-validation ensures reliable parameter selection

5. **Practical Implications**:
   - Model Selection Trade-offs:
     * Decision Trees: Better interpretability, clear decision rules
     * Random Forest: Higher accuracy, robust predictions
   - Use Case Considerations:
     * For explanatory analysis: Decision tree might be preferred
     * For predictive accuracy: Tuned random forest is recommended

6. **Recommendations**:
   - Choose model based on specific needs (interpretability vs accuracy)
   - Consider computational resources (simpler models for limited resources)
   - Regular model retraining may be needed for real-world applications
   - Monitor performance across all classes, not just overall accuracy
