---
title: "Decision Tree and Random Forest Models Comparison"
author: "Roll Number: 17"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: default
    slide-number: true
    code-fold: true
    incremental: true
    highlight-style: github
---

## Overview

- Data Generation and Preparation
- Model Fitting and Evaluation
- Model Comparison
- Conclusions

## Data Generation

### Required Packages

```{r setup, message=FALSE}
# Load required packages
library(rpart)        # For decision trees
library(randomForest) # For random forest
library(caret)        # For model evaluation
library(ggplot2)      # For visualization
```

### Data Generation Code

```{r data_generation}
# Set random seed using roll number
set.seed(17)

# Generate 4 continuous predictor variables
x1 <- rnorm(500, mean = 0, sd = 1)
x2 <- rnorm(500, mean = 2, sd = 2)
x3 <- rnorm(500, mean = -1, sd = 1.5)
x4 <- rnorm(500, mean = 1, sd = 1)

# Generate categorical response variable
y <- ifelse(x1 + x2 + rnorm(500, 0, 0.5) > 2, 1,
     ifelse(x3 - x4 + rnorm(500, 0, 0.5) < 0, 2, 3))
y <- factor(y, levels = c(1, 2, 3))

# Create and split data
data <- data.frame(x1, x2, x3, x4, result = y)
set.seed(17)
ind <- sample(2, nrow(data), prob=c(0.7,0.3), replace = T)
train_data <- data[ind==1, ]
test_data <- data[ind==2, ]
```

## Model Fitting

### Decision Tree Model

```{r decision_tree}
# Fit and plot decision tree
tree_model <- rpart(result ~ ., data = train_data, method = "class")
par(mar = c(5.1, 4.1, 6.1, 2.1))
plot(tree_model, uniform = TRUE, main = "Decision Tree")
text(tree_model, use.n = TRUE, all = TRUE, cex = 0.8)
```

#### Interpretation: Decision Tree Plot

- The plot above visualizes the structure of the fitted decision tree.
- Each node shows a split based on a predictor variable (e.g., x1, x2, x3, x4).
- The numbers at each node represent the number of observations for each class after the split.
- The tree helps us understand the sequence of decisions used to classify the data.
- Shallower splits (closer to the root) indicate more important variables for the initial classification.
- This plot is useful for model interpretability and understanding how predictions are made.

### Random Forest Model

```{r random_forest}
# Fit random forest
rf_model <- randomForest(result ~ ., data = train_data, ntree = 500)

# Plot variable importance
varImpPlot(rf_model, main = "Variable Importance Plot")
```

#### Interpretation: Variable Importance Plot

- The plot above shows the importance of each predictor variable in the random forest model.
- The x-axis (MeanDecreaseGini) measures how much each variable contributes to reducing impurity in the forest.
- Variables with higher values (e.g., x2) are more important for accurate classification.
- This plot helps identify which features are most influential in the model's predictions.
- In this example, x2 is the most important variable, followed by x1, x3, and x4.
```

## Model Evaluation

### Confusion Matrices

```{r evaluation}
# Decision Tree predictions
tree_pred <- predict(tree_model, test_data, type = "class")
tree_cm <- confusionMatrix(tree_pred, test_data$result)

# Random Forest predictions
rf_pred <- predict(rf_model, test_data)
rf_cm <- confusionMatrix(rf_pred, test_data$result)

# Print confusion matrices
print("Decision Tree Confusion Matrix:")
print(tree_cm$table)
print("\nRandom Forest Confusion Matrix:")
print(rf_cm$table)
```

## Model Comparison

### Performance Metrics

```{r comparison}
# Create comparison table
comparison <- data.frame(
  Model = c("Decision Tree", "Random Forest"),
  Accuracy = c(tree_cm$overall["Accuracy"], rf_cm$overall["Accuracy"]),
  Kappa = c(tree_cm$overall["Kappa"], rf_cm$overall["Kappa"])
)

# Display comparison
knitr::kable(comparison, digits = 4)
```

### Key Findings

- **Accuracy Comparison**:
  - Decision Tree: 84.56%
  - Random Forest: 87.25%

- **Kappa Statistics**:
  - Decision Tree: 0.7010
  - Random Forest: 0.7574

## Model Selection

### Decision Tree Advantages
- Simpler, more interpretable model
- Clear decision rules
- Lower computational cost

### Random Forest Advantages
- Higher accuracy (87.25% vs 84.56%)
- Better Kappa statistic (0.7574 vs 0.7010)
- More robust predictions
- Built-in feature importance

## Conclusion

- Random Forest performs better in terms of both accuracy and Kappa statistic
- The improvement in accuracy (2.69%) and Kappa (0.0564) suggests Random Forest is the better choice
- Variable importance plot shows which predictors are most influential
- Consider using Random Forest when prediction accuracy is the primary goal

## Thank You!

Questions? 