---
title: "Project 3 - Part 2"
author: "Mandip Chaudhary (17)"
date: "2025-04-16"
output:
  html_document: 
    toc: true
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

[Previous sections remain unchanged...]

### 7.3 K-core Decomposition

```{r kcore}
# Calculate coreness
kc <- coreness(net, mode="all")

# Visualize k-cores
plot(net, 
     vertex.size=kc*6, 
     vertex.label=kc, 
     vertex.color=colrs[kc],
     main="K-core Decomposition")

# Print coreness distribution
cat("Distribution of coreness values:\n")
print(table(kc))

# Identify nodes in the highest k-core
max_core <- max(kc)
core_nodes <- V(net)$media[kc == max_core]
cat("\nNodes in highest k-core (k =", max_core, "):\n")
print(core_nodes)
```

**Interpretation:**
K-core analysis reveals:
- The hierarchical structure of the network
- Core-periphery patterns in media relationships
- The most densely connected subgroups
- The nestedness of network communities

## 8. Assortativity and Homophily

```{r assortativity}
# Calculate different types of assortativity
type_assort <- assortativity_nominal(net, V(net)$media.type, directed=F)
size_assort <- assortativity(net, V(net)$audience.size, directed=F)
degree_assort <- assortativity_degree(net, directed=F)

# Print results
cat("Assortativity coefficients:\n")
cat("By media type:", round(type_assort, 3), "\n")
cat("By audience size:", round(size_assort, 3), "\n")
cat("By degree:", round(degree_assort, 3), "\n")

# Visualize relationships between connected nodes
# Create a scatterplot of connected nodes' audience sizes
edge_list <- as_edgelist(net)
from_size <- V(net)$audience.size[edge_list[,1]]
to_size <- V(net)$audience.size[edge_list[,2]]

plot(from_size, to_size,
     xlab="Source Node Audience Size",
     ylab="Target Node Audience Size",
     main="Audience Sizes of Connected Nodes",
     pch=19, col="orange")
```

**Interpretation:**
Assortativity analysis reveals:
- Tendency for media outlets to connect to similar/different types
- Relationship between audience size and connection patterns
- Whether high-degree nodes tend to connect to other high-degree nodes
- Patterns of homophily in the media network

## 9. Conclusions

Our comprehensive Social Network Analysis of the media network reveals several key insights:

### 9.1 Network Structure
```{r network_summary}
# Calculate key network metrics
cat("Network Summary:\n")
cat("Number of nodes:", vcount(net), "\n")
cat("Number of edges:", ecount(net), "\n")
cat("Density:", edge_density(net, loops=F), "\n")
cat("Average path length:", mean_distance(net, directed=T), "\n")
cat("Global clustering coefficient:", transitivity(net, type="global"), "\n")
```

### 9.2 Key Findings

1. **Network Topology**:
   - The network shows a moderately sparse structure (density ≈ 0.18)
   - Most media outlets are connected through short paths
   - Clear community structure exists with distinct groups

2. **Information Flow**:
   - Identified key bridging nodes through betweenness centrality
   - Hub and authority analysis revealed specialized roles
   - Reciprocal relationships indicate mutual influence

3. **Media Relationships**:
   - Found evidence of homophily based on media type
   - Audience size shows interesting patterns of connection
   - Community structure partially aligns with media categories

4. **Structural Patterns**:
   - Clear core-periphery structure identified through k-core analysis
   - Multiple communities detected through various algorithms
   - Hierarchical organization of media relationships

5. **Methodological Insights**:
   - Different layout algorithms reveal different network aspects
   - Multiple centrality measures provide complementary views
   - Various community detection methods offer different perspectives

### 9.3 Implications

This analysis helps understand:
- How information flows through the media landscape
- Which outlets serve as key information brokers
- Natural groupings and hierarchies in media relationships
- Patterns of influence and authority in media networks

The findings can inform:
- Strategic communication planning
- Understanding information dissemination patterns
- Identifying key influencers and gatekeepers
- Recognition of potential information bottlenecks

These insights contribute to our understanding of media ecosystem dynamics and can guide decisions about information distribution strategies and media engagement approaches.
