---
title: "R Notebook"
author: "Mandip Chaudhary (17)"
output:
  pdf_document:
    fig_width: 8
    fig_height: 6
    fig_caption: true
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: true
  html_notebook: default
---

7. Do the following in R Studio using "airquality" dataset of R with R script to knit PDF output.
a. Perform test of normality of "Temp" variable by each category of "Month" variable and interpret all the results carefully
```{r}
# Load necessary libraries
library(car)  # For Levene's test
library(rstatix)  # For statistical tests
library(ggplot2)  # For visualization

# Load the dataset
data(airquality)

# Convert Month to factor for better analysis
airquality$Month <- factor(airquality$Month, 
                          levels = 5:9,
                          labels = c("May", "June", "July", "August", "September"))

# a. Test of normality for each month
# Shapiro-Wilk test for each month
normality_tests <- airquality %>%
  group_by(Month) %>%
  shapiro_test(Temp)

print(normality_tests)

# Visual check of normality using Q-Q plots
ggplot(airquality, aes(sample = Temp)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~Month) +
  labs(title = "Q-Q Plots of Temperature by Month")
```
### a. Normality Test Results:
The Shapiro-Wilk test was performed for each month to test the normality of temperature distributions. A p-value > 0.05 indicates normal distribution. Based on the results, we can see that each month's P values is greater than 0.05. Thus, it follows a normal distribution for each month.


b. Perform test of equality of variance on "Temp" variable by each category of Month variable and interpret the result carefully
```{r}
# b. Test of equality of variance
# Levene's test
levene_test <- leveneTest(Temp ~ Month, data = airquality)
print("\nLevene's Test for Equality of Variance:")
print(levene_test)
```

### b. Equality of Variance Results:
The Levene's test result (p-value = 0.03941 < 0.05) indicates that the temperature variances are significantly different across months. This means the spread of temperature values is not uniform across all months - some months show more temperature variation than others.



c. Compare Temp by Month using the best statistical test for this data and interpret the result carefully.
```{r}
# c. ANOVA test
# One-way ANOVA
anova_result <- aov(Temp ~ Month, data = airquality)
print("\nANOVA Results:")
print(summary(anova_result))
```

### c. ANOVA Results:
The one-way ANOVA test was performed to compare mean temperatures across different months. The F-statistic and p-value indicate whether there are significant differences in temperature means across months.
The ANOVA results show a significant difference in mean temperatures across months (p-value < 0.05). This suggests that at least one month's temperature is significantly different from the others. However, ANOVA does not specify which months differ, so a post-hoc test is necessary to identify specific differences.


d. Perform the post-hoc test and interpret the results carefully
```{r}
# d. Post-hoc analysis
# Tukey's HSD test
tukey_result <- TukeyHSD(anova_result)
print("\nTukey's HSD Post-hoc Test Results:")
print(tukey_result)

# Visual representation of results
ggplot(airquality, aes(x = Month, y = Temp, fill = Month)) +
  geom_boxplot() +
  labs(title = "Temperature Distribution by Month",
       x = "Month",
       y = "Temperature (°F)") +
  theme_minimal()
```

### d. Post-hoc Analysis Results:
Tukey's HSD test was performed to identify which specific months have significantly different temperatures from each other. The results show pairwise comparisons between months with adjusted p-values.

The summer months (June-August) are the hottest, with May being the coolest, and September showing a transition back to cooler temperatures.

The boxplot visualization helps in understanding the distribution of temperatures across different months, showing the median, quartiles, and potential outliers.




