---
title: "Question 9"
author: "Raju Bhujel (22)"
output:
  pdf_document: default
  html_notebook: default
---

9. Do as follows using in-built "USArrests" dataset with R script to knit PDF output.

a. Create a "criminality scale" of four variables of this dataset using the Principal Component Analysis
```{r}
# Load required libraries
library(psych)
library(ggplot2)

# Perform PCA on USArrests dataset
pca_result <- prcomp(USArrests, scale = TRUE)

# Display summary of PCA results
summary(pca_result)
```

b. Compute the eigenvalues and interpret the PCA result carefully using Kaiser's criteria
```{r}
# Extract eigenvalues from PCA results
eigenvalues <- pca_result$sdev^2

# Create a data frame for eigenvalues
eigen_df <- data.frame(
  Component = 1:length(eigenvalues),
  Eigenvalue = eigenvalues
)

# Display eigenvalues
print("Eigenvalues:")
print(eigenvalues)

# Kaiser's criterion (eigenvalues > 1)
print("Components meeting Kaiser's criterion (eigenvalue > 1):")
print(sum(eigenvalues > 1))
```

c. Show the Scree plot and decide the number of components to retain with careful interpretation
```{r}
# Create Scree plot
ggplot(eigen_df, aes(x = Component, y = Eigenvalue)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(title = "Scree Plot",
       x = "Component Number",
       y = "Eigenvalue") +
  theme_minimal()
```

d. Revise the criminality scale using varimax rotation and interpret the results carefully
```{r}
# Perform PCA with varimax rotation
pca_rotated <- principal(USArrests, nfactors = 2, rotate = "varimax")

# Display rotated component loadings
print("Rotated Component Loadings:")
print(pca_rotated$loadings, cutoff = 0.3)

# Display variance explained by rotated components
print("Variance Explained by Rotated Components:")
print(pca_rotated$Vaccounted)
```