---
title: "Question 10"
author: "Mandip Chaudhary (17)"
output:
  pdf_document: default
---

10. Use the four variables of "USArrests" data file into R Studio and do as follows with R script to knit PDF output.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(factoextra)
```

a. Fit a k-means clustering model in the data with k = 2, plot it in the single graph and interpret it carefully

```{r kmeans_2}
# Load the USArrests dataset
data("USArrests")

# Scale the data
scaled_data <- scale(USArrests)

# Perform k-means clustering with k=2
set.seed(123)  # for reproducibility
km2 <- kmeans(scaled_data, centers = 2, nstart = 25)

# Plot the clusters
fviz_cluster(km2, data = scaled_data,
             palette = c("#2E9FDF", "#FC4E07"),
             geom = "point",
             ellipse.type = "convex",
             ggtheme = theme_minimal(),
             main = "K-means Clustering with K=2")
```

**Interpretation for k=2:**
The k-means clustering with k=2 divides the US states into two distinct groups based on their crime rates. The plot shows a clear separation between the clusters, where:
- Cluster 1 (blue) represents states with generally lower crime rates
- Cluster 2 (red) represents states with generally higher crime rates
The separation suggests a natural dichotomy in crime patterns across US states.

b. Add cluster centers for the plot of clusters formed with k = 2 and interpret it carefully

```{r kmeans_2_centers}
# Plot clusters with centers
fviz_cluster(km2, data = scaled_data,
             palette = c("#2E9FDF", "#FC4E07"),
             geom = c("point", "text"),
             ellipse.type = "convex",
             ggtheme = theme_minimal(),
             main = "K-means Clustering with K=2 (Including Centers)")

# Print cluster centers
print("Cluster Centers:")
print(km2$centers)
```

**Interpretation of Cluster Centers for k=2:**
The cluster centers represent the average values of the four variables (Murder, Assault, UrbanPop, and Rape) for each cluster. These centers help us understand the characteristics of each cluster:
- The cluster centers show the average crime rates and urban population percentage for each group
- The distance between centers indicates how distinct the clusters are
- States closer to a center are more typical of that cluster's characteristics

c. Fit a k-means clustering model in the data with k = 3, plot it in the single graph and interpret it carefully

```{r kmeans_3}
# Perform k-means clustering with k=3
set.seed(123)
km3 <- kmeans(scaled_data, centers = 3, nstart = 25)

# Plot the clusters
fviz_cluster(km3, data = scaled_data,
             palette = c("#2E9FDF", "#FC4E07", "#00AFBB"),
             geom = "point",
             ellipse.type = "convex",
             ggtheme = theme_minimal(),
             main = "K-means Clustering with K=3")
```

**Interpretation for k=3:**
The k-means clustering with k=3 provides a more nuanced view of crime patterns across US states:
- Cluster 1 (blue) represents states with moderate crime rates
- Cluster 2 (red) represents states with high crime rates
- Cluster 3 (green) represents states with low crime rates
This three-cluster solution reveals more detailed patterns in the data compared to the two-cluster solution.

d. Add cluster centers for the plot of clusters formed with k = 3 and interpret it carefully

```{r kmeans_3_centers}
# Plot clusters with centers
fviz_cluster(km3, data = scaled_data,
             palette = c("#2E9FDF", "#FC4E07", "#00AFBB"),
             geom = c("point", "text"),
             ellipse.type = "convex",
             ggtheme = theme_minimal(),
             main = "K-means Clustering with K=3 (Including Centers)")

# Print cluster centers
print("Cluster Centers:")
print(km3$centers)
```

**Interpretation of Cluster Centers for k=3:**
The three cluster centers provide more detailed insights into the crime patterns:
- Each center represents the average characteristics of its cluster
- The relative positions of the centers show how the clusters differ in terms of crime rates and urban population
- The distances between centers indicate the distinctness of each cluster
- This three-cluster solution provides a more granular view of crime patterns across US states compared to the two-cluster solution